Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-02-03 paper/zhang-cook-laa-langrene-menendez.Rmd
Updated text in the first 2 sections

Visualisation is widely used in exploratory data analysis [@tukey1977exploratory; @unwin2015graphical;  @healy2018data; @wilke2019fundamentals]. Presenting information in graphics often unveils information that would otherwise not be aware of and provides a more comprehensive understanding of the problem at hand. Task specific tools such as @li2020visualizing show how visualisation can be used to understand the behaviour of neural network on classification models, but no general visualisation tool is available for diagnosing optimisation procedures. The work presented in this paper brings visualization tools into optimisation problems with an aim to better understand the performance of the optimisers in practice.

@@

Projection pursuit guided tour combines two different methods (projection pursuit and guided tour) in exploratory data analysis. Projection pursuit, coined by @friedman1974projection, detects interesting structures (e.g. clustering, outliers and skewness) in multivariate data via low dimensions projection. Guided tour is one variation of a broader class of data visualisation methods, tour, which visualises high-dimensional data through a series of animated projection.

Notation is first established before defining projection pursuit. Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A d-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$, and defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis Define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto an scalar. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. A number of index functions have been proposed in the literature to detect different data structures [friedman1974projection; @hall1989polynomial; @cook1993projection; @posse1995projection] and to classify groups [@lee2005projection; @lee2010projection].

As a general visualisation method, tour produces animations of high dimensional data via rotations low dimension planes. Different tour types choose these planes differently: grand tour [@cook2008grand] selects the planes randomly to provide a general overview;  manual tour [@cook1997manual] gradually phases in and out one variable to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses the planes with the aid of projection pursuit, to gradually reveal the most interesting projection. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs the geodesic interpolation between these planes to form a tour path. Figure \ref{fig:tour-path} shows a sketch of the tour path where the blue frames are produced by the projection pursuit optimisation and the white frames are interpolations  between the blue frames. Mathematical details of the geodesic interpolation can be found in @buja2005computational. The tour method has been implemented in the R package \CRANpkg{tourr} [@tourr].

>>>>

Visualisation is widely used in exploratory data analysis [@tukey1977exploratory; @unwin2015graphical;  @healy2018data; @wilke2019fundamentals]. Presenting information in graphics often unveils information that would otherwise not be discovered and provides a more comprehensive understanding of the problem at hand. Task specific tools such as @li2020visualizing show how visualisation can be used to understand, for instance, the behaviour of neural network on classification models. However, no general visualisation tool is available for diagnosing optimisation procedures. The work presented in this paper brings visualization tools into optimisation problems with the aim to better understand the performance of optimisers in practice.

@@

Projection pursuit guided tour combines two different methods (projection pursuit and guided tour) in exploratory data analysis. Projection pursuit, coined by @friedman1974projection, detects interesting structures (e.g. clustering, outliers and skewness) in multivariate data via low dimensions projection. Guided tour is one variation of a broader class of data visualisation methods, tour, which displays high-dimensional data through a series of animated projections.

<!-- Notation is first established before defining projection pursuit. -->
Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A d-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto an scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. A number of index functions have been proposed in the literature to detect different data structures [friedman1974projection; @hall1989polynomial; @cook1993projection; @posse1995projection] and to classify groups [@lee2005projection; @lee2010projection].

As a general visualisation method, tour produces animations of high dimensional data via rotations of low dimension planes. There are different tour implementations depending on how the high dimensional space is investigated: grand tour [@cook2008grand] selects the planes randomly to provide a general overview;  manual tour [@cook1997manual] gradually phases in and out one variable to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses the planes with the aid of projection pursuit, to gradually reveal the most interesting projection. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs a geodesic interpolation between these planes to form a tour path. Figure \ref{fig:tour-path} shows a sketch of the tour path where the blue frames are produced by the projection pursuit optimisation and the white frames are interpolations  between the blue frames. Mathematical details of the geodesic interpolation can be found in @buja2005computational. The aforementioned tour method has been implemented in the R package \CRANpkg{tourr} [@tourr].

### 2021-02-03 paper/zhang-cook-laa-langrene-menendez.tex
underscore to dash in tour-path.png
Updated text in the first 2 sections

Visualisation is widely used in exploratory data analysis
\citep{tukey1977exploratory, unwin2015graphical, healy2018data, wilke2019fundamentals}.
Presenting information in graphics often unveils information that would
otherwise not be aware of and provides a more comprehensive
understanding of the problem at hand. Task specific tools such as
\citet{li2020visualizing} show how visualisation can be used to
understand the behaviour of neural network on classification models, but
no general visualisation tool is available for diagnosing optimisation
procedures. The work presented in this paper brings visualization tools
into optimisation problems with an aim to better understand the
performance of the optimisers in practice.

@@

Projection pursuit guided tour combines two different methods
(projection pursuit and guided tour) in exploratory data analysis.
Projection pursuit, coined by \citet{friedman1974projection}, detects
interesting structures (e.g.~clustering, outliers and skewness) in
multivariate data via low dimensions projection. Guided tour is one
variation of a broader class of data visualisation methods, tour, which
visualises high-dimensional data through a series of animated
projection.

Notation is first established before defining projection pursuit. Let
\(\mathbf{X}_{n \times p}\) be the data matrix with \(n\) observations
in \(p\) dimensions. A d-dimensional projection is a linear
transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\), and defined
as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis Define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto an scalar.
This is commonly known as the projection pursuit index function, or just
index function, and is used to measure the ``interestingness'' of a
given projection. A number of index functions have been proposed in the
literature to detect different data structures
\citep[friedman1974projection;][]{hall1989polynomial, cook1993projection, posse1995projection}
and to classify groups \citep{lee2005projection, lee2010projection}.

As a general visualisation method, tour produces animations of high
dimensional data via rotations low dimension planes. Different tour
types choose these planes differently: grand tour \citep{cook2008grand}
selects the planes randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs the
geodesic interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The tour method has been implemented in
the R package \CRANpkg{tourr} \citep{tourr}.

@@

{\centering \includegraphics[width=0.5\linewidth,height=0.2\textheight]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/img/tour_path}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-search-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-interp-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-animated-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-tour-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/interruption-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/polish-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/flip-sign-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/noisy-better-geo-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-result-1}

>>>>

Visualisation is widely used in exploratory data analysis
\citep{tukey1977exploratory, unwin2015graphical, healy2018data, wilke2019fundamentals}.
Presenting information in graphics often unveils information that would
otherwise not be discovered and provides a more comprehensive
understanding of the problem at hand. Task specific tools such as
\citet{li2020visualizing} show how visualisation can be used to
understand, for instance, the behaviour of neural network on
classification models. However, no general visualisation tool is
available for diagnosing optimisation procedures. The work presented in
this paper brings visualization tools into optimisation problems with
the aim to better understand the performance of optimisers in practice.

@@

Projection pursuit guided tour combines two different methods
(projection pursuit and guided tour) in exploratory data analysis.
Projection pursuit, coined by \citet{friedman1974projection}, detects
interesting structures (e.g.~clustering, outliers and skewness) in
multivariate data via low dimensions projection. Guided tour is one
variation of a broader class of data visualisation methods, tour, which
displays high-dimensional data through a series of animated projections.

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A d-dimensional projection is a linear
transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\) defined as
\(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto an scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. A number of index functions have been proposed in
the literature to detect different data structures
\citep[friedman1974projection;][]{hall1989polynomial, cook1993projection, posse1995projection}
and to classify groups \citep{lee2005projection, lee2010projection}.

As a general visualisation method, tour produces animations of high
dimensional data via rotations of low dimension planes. There are
different tour implementations depending on how the high dimensional
space is investigated: grand tour \citep{cook2008grand} selects the
planes randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs a geodesic
interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The aforementioned tour method has been
implemented in the R package \CRANpkg{tourr} \citep{tourr}.

@@

{\centering \includegraphics[width=0.5\linewidth,height=0.2\textheight]{/Users/pmen0008/Documents/paper-tour-vis/img/tour-path}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/toy-search-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/toy-interp-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/toy-pca-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/toy-pca-animated-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/toy-tour-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/interruption-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/polish-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/flip-sign-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/noisy-better-geo-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/kol-result-1}
